Continous time Markov chain models; \\ extension, simulation and visualisation

### 2015-02-24 chapters/chapter01.tex
fixed typo in display 1.13
preparing manuscript for printing

%likelihood
As a statistical alternative to an optimality criterion, the likelihood had also been proposed for phylogenetic inference relatively early.
The idea of using maximum likelihood to estimate evolutionary trees was first proposed by \cite{Edwards1964}, although without a working implementation at the time of the publication.
Maximum likelihood is a probabilistic method of inference, that evaluates a hypothesis in terms of how probable it is in the light of the observed data.
In a phylogenetic setting, the likelihood function returns the probability that the proposed model and the proposed tree, understood as a topology and branch lengths, have generated the observed data.
\myedit{svthree}{
In general an evolutionary history and associated set of evolutionary parameters with higher probability of generating the observed data is naturally preferred over a history with a lower probability of generating it.
By using optimization techniques one can find the parameter values that maximize the probability of observing the data.
}
% The evolutionary history and associated set of evolutionary parameters with a higher probability is naturally preferred over a history with a lower probability.
% By using optimization techniques one can find the parameter values that maximize the probability of observing the data.
In his influential work, \cite{Felsenstein1981} proposed an efficient algorithm for calculating a likelihood of sequence evolution on a tree under an evolutionary model of substitution, the so called `tree pruning' algorithm.
Unfortunately finding the maximum likelihood tree requires a search through the space of all possible tree topologies; in section \ref{sub:mlburden} we discuss in detail the computational burden associated with these calculations.
The effort required to exhaustively search through a space of all possible rooted trees grows explosively with the number of taxa, as depicted in Table \ref{tab:maxLikeBurden}.

@@

%gpus
\myedit{DPfive}{
% The growing availability of large complex data-sets has led to the term `Big Data'. %being coined.
% Large sample sizes, such as full genome sequences or the ones generated with next-generation sequencing techniques are increasingly challenging phylogenetic inference.
The growing availability of large and complex data-sets, such as full genome sequences or the ones generated with next-generation sequencing techniques, are increasingly challenging phylogenetic inference.
}%END: myedit
Fortunately strategies have been proposed to reduce the time that it takes to fit complex phylogenetic models to vast amounts of data.
These solutions are mainly focused on parallellization and on harnessing the capabilities of modern graphics processing units (GPUs), such as the NVIDIAs Tesla line of GPUs that target the high performance computing market.
BEAGLE (Broad-platform evolutionary analysis generalized likelihood evaluator, \cite{Ayres2012}), is a library that can be used in conjunction with BEAST and provides an API (Application Programming Interface) that massively parallel graphics processors and other multicore devices can exploit to substantially decrease the runtime of phylogenetic analysis \citep{Suchard2009}.
%BEAST readily integrates with BEAGLE and provides a GUI interface to configure the library.

@@

\begin{equation}
\left\{ \left\{ P(t)\right\} _{ij} \right\} _{ij} = \left\{ e^{\mathbf{R}r t}e^{-r t} \right\} _{ij} = \left\{ e^{(\mathbf{R}-\mathbf{I})r t} \right\} _{ij}= \left\{ e^{\mathbf{Q}r t} \right\} _{ij}
\label{eq:matrixExp2}
\end{equation}

@@

As $t$ goes to infinity the probability that the site is in some state $j$ is non-zero and does not depend on the starting state $i$.
This implies that the chain will always display the same long-term behavior, which is why $\mathbf{\Pi}=\{\pi_{i},\ i\in\mathcal{E}\}$ is called a \emph{stationary distribution} (also referred to as the \emph{equilibrium distribution} or \emph{equilibrium frequencies}).
We can look at the elements of $\mathbf{\Pi}$ as the proportion of time spent in state $j\in\mathcal{E}$ after the Markov chain has run for infinitely long time, that does not depend upon the initial state $i$ of the process.
In other words, if we sample the initial state from the stationary distribution and then let the process run for some sufficiently long time $t$, then the distribution of the final state will equal the stationary distribution. 
\myedit{svsixone}{
From the biological perspective stationarity means that the probability distribution of a certain state (e.g. nucleotide character) is constant over the whole course of evolution; the probability to draw a certain state is the same whenever the sampling is done.
A process can be stationary and not be homogeneous, for example by varying substitution rates on the tree branches (see \cite{Yang2002} and the discussion in Subsection~\ref{sub:dpp}).
However the converse is not true, and non-stationarity induces non-homogeneity, as the stochastic process of evolution depends on the stationary distribution.
}

@@

\myedit{svsixtwo}{
Realistic substitution processes are however unlikely to be time-reversible, especially since abandoning stationarity and homogeneity assumptions implies abandoning reversibility of the evolutionary process.
Recently however procedures have also been developed to compute phylogenetic likelihoods under non-reversible models \citep{Boussau2006}, and such non-reversible models have also been explored for large state-space discrete phylogeographic applications \citep{Edwards2011}.
}

@@

\myedit{svonea}{
First two computationally tractable codon substitution models were independently proposed by \cite{Muse1994} and \cite{Goldman1994} in the same issue of Molecular Biology and Evolution (MBE).
In these models, a non-stop codon triplet $n_{1}n_{2}n_{3}$ is considered to be the smallest unit of evolution.
According to the universal genetic code, there are $4^3$ possible triplets minus three stop codons, resulting in a state space size of $61$ codons.
The models inherit the standard assumption of independence, i.e. the substitutions at these three codon positions occur independently, and only a single change per triplet can occur at a given time. 
%%GB: explain in more detail, i.e. more than 1 change is possible in subsequent infinitesimal time intervals
Proteins are coded by a set of 20 amino acids, with each amino acid being coded by a codon triplet. 
Because there are 61 non-stop codons to encode for 20 amino acids, some amino acids will inevitably be coded by more than one codon.
A substitution that does not change the encoding amino-acid is called a \emph{synonymous} substitution, while a \emph{non-synonymous} substitution refers to a substitution that does result in an amino acid substitution.
Quantifying the rate at which these substitutions occur allows us to asses the nature of selective pressure, which is the main driving force behind molecular evolution.
}

@@

\myedit{svoneb}{
We start by considering heights of tree tips and introducing a new parameter, the rate of mutation $r$, which here for simplicity is assumed to apply to the whole tree (a strict clock), which allows us to change the scale of the tree from units of substitutions to a time-scale measured in years.

Both the rate of evolution $r$ and the unobserved heights $T_{0},T_{6},T_{7},T_{8}$ are estimated from the molecular sequence data, with the constraint that the heights are proportional to their sampling time and the ratio of this proportionality being $r$.
}

@@

\noindent
where $\mathbf{\Theta}$ represents the tree topology, the branch lengths, the parameters of the substitution model and other parameters of interest that we want to estimate.
This quantity expresses how likely it is to observe the data given a specific value of the parameter(s) of interest $\mathbf{\Theta}$.
\myedit{sveight}{
By finding the value of $\mathbf{\Theta}$ that maximizes (\ref{eq:likelihood}) we obtain an estimate of the most plausible 
%hypothesis for the data.
set of parameters given the observed data.}

@@

\myedit{svnine}{
The prior choice is perhaps one of the most controversial aspects of the Bayesian inference.
The prior distributions may be inspired by previous analyses of other sets of data, or they may form some set of biologically plausible values.
However in some cases there may be no suitable objective prior knowledge at hand, in which case one often resorts to the so-called \emph{vague} or \emph{diffuse} priors, which have large variances.
Some popular uninformative priors on continuous, unbounded variables may be improper, \textit{i.e.} their distribution is infinitesimal on an infinite range in order to integrate to $1$.
Among examples of such priors is the uniform distribution on an infinite interval.
Although many of these prior choices still lead to a proper posterior distribution, some controversies about their choice may still arise, in particular when the data is poorly informative.
}

@@

\myedit{sveleven}{
Figure~\ref{fig:bayes1} shows the resulting posterior distribution along with the prior and the scaled likelihood.
The likelihood of the binomial model used in this example is simply:
}

@@

\myedit{svonec}{
Central to the Bayesian phylogeographic framework is the hypothesis that sequence evolution is occurring simultaneously with the process of geographical dispersal. 
The observed data, residing at the $N$ tips of phylogeny $\mathbf{F}$, is recorded as both character sequence data $\mathbf{X}=(X_{1},...,X_{N})$ and discretized geographic locations (countries, districts, cities) $\mathbf{Y}=(Y_{1},...,Y_{N})$, as depicted in the upper facet of Figure~\ref{fig:discrete_phylo_illust}.
We assume independent stochastic processes are responsible for generating these data, with ancestral states of $\mathbf{X}$ generated as before by a CTMC characterized by rate matrix $\mathbf{Q}$ and the unobserved ancestral locations $(Y_{N+1},...,Y_{2N-2})$ generated by a separate CTMC characterized by rate matrix $\mathbf{\Phi}$ (see Figure~\ref{fig:discrete_phylo_illust}, lower facet).
The entries of this matrix are the transition rates $\phi_{ij}$ between locations $i$ and $j$.
Unlike before, the rate matrix does not necessarily have to be symmetrical, i.e. there exist some $i,j$ such that $\phi_{ij}\neq\phi_{ji}$.
Additionally, a priori one would expect many of the infinitesimal rates to be zero. 
}

>>>>

%likelihood
As a statistical alternative to an optimality criterion, the likelihood had also been proposed for phylogenetic inference relatively early.
The idea of using maximum likelihood to estimate evolutionary trees was first proposed by \cite{Edwards1964}, although without a working implementation at the time of the publication.
Maximum likelihood is a probabilistic method of inference, that evaluates a hypothesis in terms of how probable it is in the light of the observed data.
In a phylogenetic setting, the likelihood function returns the probability that the proposed model and the proposed tree, understood as a topology and branch lengths, have generated the observed data.
% \myedit{svthree}{
In general an evolutionary history and associated set of evolutionary parameters with higher probability of generating the observed data is naturally preferred over a history with a lower probability of generating it.
By using optimization techniques one can find the parameter values that maximize the probability of observing the data.
% }
% The evolutionary history and associated set of evolutionary parameters with a higher probability is naturally preferred over a history with a lower probability.
% By using optimization techniques one can find the parameter values that maximize the probability of observing the data.
In his influential work, \cite{Felsenstein1981} proposed an efficient algorithm for calculating a likelihood of sequence evolution on a tree under an evolutionary model of substitution, the so called `tree pruning' algorithm.
Unfortunately finding the maximum likelihood tree requires a search through the space of all possible tree topologies; in section \ref{sub:mlburden} we discuss in detail the computational burden associated with these calculations.
The effort required to exhaustively search through a space of all possible rooted trees grows explosively with the number of taxa, as depicted in Table \ref{tab:maxLikeBurden}.

@@

%gpus
% \myedit{DPfive}{
% The growing availability of large complex data-sets has led to the term `Big Data'. %being coined.
% Large sample sizes, such as full genome sequences or the ones generated with next-generation sequencing techniques are increasingly challenging phylogenetic inference.
The growing availability of large and complex data-sets, such as full genome sequences or the ones generated with next-generation sequencing techniques, are increasingly challenging phylogenetic inference.
% }%END: myedit
Fortunately strategies have been proposed to reduce the time that it takes to fit complex phylogenetic models to vast amounts of data.
These solutions are mainly focused on parallellization and on harnessing the capabilities of modern graphics processing units (GPUs), such as the NVIDIAs Tesla line of GPUs that target the high performance computing market.
BEAGLE (Broad-platform evolutionary analysis generalized likelihood evaluator, \cite{Ayres2012}), is a library that can be used in conjunction with BEAST and provides an API (Application Programming Interface) that massively parallel graphics processors and other multicore devices can exploit to substantially decrease the runtime of phylogenetic analysis \citep{Suchard2009}.
%BEAST readily integrates with BEAGLE and provides a GUI interface to configure the library.

@@

\begin{equation}
\left\{ P(t)\right\} _{ij}  = \left\{ e^{\mathbf{R}r t}e^{-r t} \right\} _{ij} = \left\{ e^{(\mathbf{R}-\mathbf{I})r t} \right\} _{ij}= \left\{ e^{\mathbf{Q}r t} \right\} _{ij}
\label{eq:matrixExp2}
\end{equation}

@@

As $t$ goes to infinity the probability that the site is in some state $j$ is non-zero and does not depend on the starting state $i$.
This implies that the chain will always display the same long-term behavior, which is why $\mathbf{\Pi}=\{\pi_{i},\ i\in\mathcal{E}\}$ is called a \emph{stationary distribution} (also referred to as the \emph{equilibrium distribution} or \emph{equilibrium frequencies}).
We can look at the elements of $\mathbf{\Pi}$ as the proportion of time spent in state $j\in\mathcal{E}$ after the Markov chain has run for infinitely long time, that does not depend upon the initial state $i$ of the process.
In other words, if we sample the initial state from the stationary distribution and then let the process run for some sufficiently long time $t$, then the distribution of the final state will equal the stationary distribution. 
% \myedit{svsixone}{
From the biological perspective stationarity means that the probability distribution of a certain state (e.g. nucleotide character) is constant over the whole course of evolution; the probability to draw a certain state is the same whenever the sampling is done.
A process can be stationary and not be homogeneous, for example by varying substitution rates on the tree branches (see \cite{Yang2002} and the discussion in Subsection~\ref{sub:dpp}).
However the converse is not true, and non-stationarity induces non-homogeneity, as the stochastic process of evolution depends on the stationary distribution.
% }

@@

% \myedit{svsixtwo}{
Realistic substitution processes are however unlikely to be time-reversible, especially since abandoning stationarity and homogeneity assumptions implies abandoning reversibility of the evolutionary process.
Recently however procedures have also been developed to compute phylogenetic likelihoods under non-reversible models \citep{Boussau2006}, and such non-reversible models have also been explored for large state-space discrete phylogeographic applications \citep{Edwards2011}.
% }

@@

% \myedit{svonea}{
First two computationally tractable codon substitution models were independently proposed by \cite{Muse1994} and \cite{Goldman1994} in the same issue of Molecular Biology and Evolution (MBE).
In these models, a non-stop codon triplet $n_{1}n_{2}n_{3}$ is considered to be the smallest unit of evolution.
According to the universal genetic code, there are $4^3$ possible triplets minus three stop codons, resulting in a state space size of $61$ codons.
The models inherit the standard assumption of independence, i.e. the substitutions at these three codon positions occur independently, and only a single change per triplet can occur at a given time. 
%%GB: explain in more detail, i.e. more than 1 change is possible in subsequent infinitesimal time intervals
Proteins are coded by a set of 20 amino acids, with each amino acid being coded by a codon triplet. 
Because there are 61 non-stop codons to encode for 20 amino acids, some amino acids will inevitably be coded by more than one codon.
A substitution that does not change the encoding amino-acid is called a \emph{synonymous} substitution, while a \emph{non-synonymous} substitution refers to a substitution that does result in an amino acid substitution.
Quantifying the rate at which these substitutions occur allows us to asses the nature of selective pressure, which is the main driving force behind molecular evolution.
% }

@@

% \myedit{svoneb}{
We start by considering heights of tree tips and introducing a new parameter, the rate of mutation $r$, which here for simplicity is assumed to apply to the whole tree (a strict clock), which allows us to change the scale of the tree from units of substitutions to a time-scale measured in years.

Both the rate of evolution $r$ and the unobserved heights $T_{0},T_{6},T_{7},T_{8}$ are estimated from the molecular sequence data, with the constraint that the heights are proportional to their sampling time and the ratio of this proportionality being $r$.
% }

@@

\noindent
where $\mathbf{\Theta}$ represents the tree topology, the branch lengths, the parameters of the substitution model and other parameters of interest that we want to estimate.
This quantity expresses how likely it is to observe the data given a specific value of the parameter(s) of interest $\mathbf{\Theta}$.
% \myedit{sveight}{
By finding the value of $\mathbf{\Theta}$ that maximizes (\ref{eq:likelihood}) we obtain an estimate of the most plausible 
%hypothesis for the data.
set of parameters given the observed data.
% }

@@

% \myedit{svnine}{
The prior choice is perhaps one of the most controversial aspects of the Bayesian inference.
The prior distributions may be inspired by previous analyses of other sets of data, or they may form some set of biologically plausible values.
However in some cases there may be no suitable objective prior knowledge at hand, in which case one often resorts to the so-called \emph{vague} or \emph{diffuse} priors, which have large variances.
Some popular uninformative priors on continuous, unbounded variables may be improper, \textit{i.e.} their distribution is infinitesimal on an infinite range in order to integrate to $1$.
Among examples of such priors is the uniform distribution on an infinite interval.
Although many of these prior choices still lead to a proper posterior distribution, some controversies about their choice may still arise, in particular when the data is poorly informative.
% }

@@

% \myedit{sveleven}{
Figure~\ref{fig:bayes1} shows the resulting posterior distribution along with the prior and the scaled likelihood.
The likelihood of the binomial model used in this example is simply:
% }

@@

% \myedit{svonec}{
Central to the Bayesian phylogeographic framework is the hypothesis that sequence evolution is occurring simultaneously with the process of geographical dispersal. 
The observed data, residing at the $N$ tips of phylogeny $\mathbf{F}$, is recorded as both character sequence data $\mathbf{X}=(X_{1},...,X_{N})$ and discretized geographic locations (countries, districts, cities) $\mathbf{Y}=(Y_{1},...,Y_{N})$, as depicted in the upper facet of Figure~\ref{fig:discrete_phylo_illust}.
We assume independent stochastic processes are responsible for generating these data, with ancestral states of $\mathbf{X}$ generated as before by a CTMC characterized by rate matrix $\mathbf{Q}$ and the unobserved ancestral locations $(Y_{N+1},...,Y_{2N-2})$ generated by a separate CTMC characterized by rate matrix $\mathbf{\Phi}$ (see Figure~\ref{fig:discrete_phylo_illust}, lower facet).
The entries of this matrix are the transition rates $\phi_{ij}$ between locations $i$ and $j$.
Unlike before, the rate matrix does not necessarily have to be symmetrical, i.e. there exist some $i,j$ such that $\phi_{ij}\neq\phi_{ji}$.
Additionally, a priori one would expect many of the infinitesimal rates to be zero. 
% }

### 2015-02-24 chapters/chapter02.tex
preparing manuscript for printing

In recent decades, substantial work has aimed at relaxing the standard assumptions in CTMC processes, in order to uncover more complex evolutionary processes and assess their impact on phylogenetic reconstruction. 
To accommodate nonstationarity, \citet{Yang1995}, \citet{Galtier1998} and \citet{Galtier1999}, for example, have proposed models that allow branch-specific nucleotide compositions.
Although this includes general treatments involving separate composition parameters for each tree branch, large trees will inevitably lead to over-parameterized models.
To address this problem, \citet{Foster2004} has developed an approach that maps a restricted, but fixed number of nucleotide composition vectors with estimable frequencies to the tree. 
%TODO
\myedit{svfifteen}{
Set in a Bayesian framework, this approach also integrates over the tree topology and finds improved posterior support estimates for topologies in examples with compositional heterogeneity, allowing for different CTMC stationary frequencies in the evolutionary history, 
as opposed to inference under a stationary model that would suffer from attraction artifacts due to similar compositional biases. 
}
Further developments have uncoupled compositional shifts from particular nodes in the tree while estimating the total number of events of compositional drift distributed across the tree using a compound Poisson process \citep{Blanquart2006}. 
\cite{Blanquart2008} have also combined such approaches for amino acid evolution with models that take into account site-specific substitution patterns induced by protein structure and function.

@@

Figure~\ref{fig:collapsing} illustrates this action for a branch that spans a single boundary at known time $T_1$ with $\mathbf{Q}_1$ governing the process before the boundary and $\mathbf{Q}_2$ after the boundary. Letting $X(T_1) = k$ represent the augmented state of the stochastic process at the boundary, we compute

@@

In Figure~\ref{fig:2RateTree}, we illustrate the tree topology and the transition times defined for both a two-epoch and three-epoch specification.
This tree has 69 tips, is rooted and time-scaled, effectively covering a period of about 18 years. 
For each replicate data set, we simulate $1000$ nucleotide sites under the Hasegawa, Kishino and Yano (HKY) model \citep{hky85}. 
We set the substitution rates and base frequencies to values estimated from the real data set.
\myedit{svtwo}{
In all our analyses of the replicate data, as well as in the analyses of the real-world data sets in the next sections, we consider the tree as random and estimate it from the nucleotide data.
}

@@

Despite the observation that the Bayesian credible intervals for patient 1, 2, 6 and 7 estimates do not overlap, this does not provide a formal test to evaluate their differences.
Therefore, we conduct a Bayes factor (BF) test \citep{Suchard2005} that expresses the posterior odds over the prior odds that $\omega_{1} < \omega_{2}$ for the individual analyses of each patient.
% To determine the posterior odds, we note that the MCMC sample average of an indicator function that the parameter values fall within one competing model space converges to the posterior probability of that model.
\myedit{sveighteen}{
To determine the posterior odds for the hypothesis that $\omega_{1} < \omega_{2}$, we note that the MCMC sample average of an indicator function that keeps track of samples for which $\omega_{1} < \omega_{2}$ converges to the posterior probability of that hypothesis.
}
The prior odds in our case is simply 1.
The log Bayes factors listed in Table \ref{tab:shank_bf} suggest generally strong evidence for a declining selective pressure after progression, with one notable exception for patient 5.
We also provide a Bayes factor that summarizes the joint evidence for $\omega_{1} < \omega_{2}$, which suggest an overall support in favor of the immune relaxation hypothesis (log BF = 2.14), in accordance with previous findings suggesting a general decrease in non-synonymous divergence at late-stage infection \citep{williamson2005, Lemey2007}.

@@

In a second application of the epoch model, we focus on discrete diffusion processes to infer spatio-temporal history from viral gene sequences.
This type of phylogeographic inference, where the sampling locations are considered as discrete geographic traits, has gained popularity in recent years, at least partly because of a flexible and efficient Bayesian implementation that connects dispersal dynamics to sequence evolution in time-measured phylogenies \citep{Lemey2009}.
Recently \citet{Bahl2011} have applied this Bayesian inference framework to investigate the circulation dynamics of global influenza A H3N2 through time.
Since the authors were interested in capturing the heterogeneity in these dynamics over successive seasonal epidemics between 2003 and 2006, they consider discrete traits that are the product of sampling location and sampling time (epidemic season).
\myedit{svsixteen}{
This means for example that samples collected from October 2002 to February 2003 in New York are assigned to the discrete state labelled as ``New York 2003'', while sampling in the sample location during another seasons will be represented by another state \citep{Bahl2011}.
}
Not only does this discretization by sampling time seem counterintuitive for a model that emits discrete outcomes as a continuous function of time, it also considerably increases the dimensionality of the CTMC rate matrix and thus the number of parameters to inform by the sparse spatial data.

>>>>

In recent decades, substantial work has aimed at relaxing the standard assumptions in CTMC processes, in order to uncover more complex evolutionary processes and assess their impact on phylogenetic reconstruction. 
To accommodate nonstationarity, \citet{Yang1995}, \citet{Galtier1998} and \citet{Galtier1999}, for example, have proposed models that allow branch-specific nucleotide compositions.
Although this includes general treatments involving separate composition parameters for each tree branch, large trees will inevitably lead to over-parameterized models.
To address this problem, \citet{Foster2004} has developed an approach that maps a restricted, but fixed number of nucleotide composition vectors with estimable frequencies to the tree. 
%TODO
% \myedit{svfifteen}{
Set in a Bayesian framework, this approach also integrates over the tree topology and finds improved posterior support estimates for topologies in examples with compositional heterogeneity, allowing for different CTMC stationary frequencies in the evolutionary history, 
as opposed to inference under a stationary model that would suffer from attraction artifacts due to similar compositional biases. 
% }
Further developments have uncoupled compositional shifts from particular nodes in the tree while estimating the total number of events of compositional drift distributed across the tree using a compound Poisson process \citep{Blanquart2006}. 
\cite{Blanquart2008} have also combined such approaches for amino acid evolution with models that take into account site-specific substitution patterns induced by protein structure and function.

@@

% \myedit{svseventeen}{
Figure~\ref{fig:collapsing} illustrates this action for a branch that spans a single boundary at known time $T_1$ with $\mathbf{Q}_1$ governing the process before the boundary and $\mathbf{Q}_2$ after the boundary. Letting $X(T_1) = k$ represent the augmented state of the stochastic process at the boundary, we compute:
% }

@@

In Figure~\ref{fig:2RateTree}, we illustrate the tree topology and the transition times defined for both a two-epoch and three-epoch specification.
This tree has 69 tips, is rooted and time-scaled, effectively covering a period of about 18 years. 
For each replicate data set, we simulate $1000$ nucleotide sites under the Hasegawa, Kishino and Yano (HKY) model \citep{hky85}. 
We set the substitution rates and base frequencies to values estimated from the real data set.
% \myedit{svtwo}{
In all our analyses of the replicate data, as well as in the analyses of the real-world data sets in the next sections, we consider the tree as random and estimate it from the nucleotide data.
% }

@@

Despite the observation that the Bayesian credible intervals for patient 1, 2, 6 and 7 estimates do not overlap, this does not provide a formal test to evaluate their differences.
Therefore, we conduct a Bayes factor (BF) test \citep{Suchard2005} that expresses the posterior odds over the prior odds that $\omega_{1} < \omega_{2}$ for the individual analyses of each patient.
% To determine the posterior odds, we note that the MCMC sample average of an indicator function that the parameter values fall within one competing model space converges to the posterior probability of that model.
% \myedit{sveighteen}{
To determine the posterior odds for the hypothesis that $\omega_{1} < \omega_{2}$, we note that the MCMC sample average of an indicator function that keeps track of samples for which $\omega_{1} < \omega_{2}$ converges to the posterior probability of that hypothesis.
% }
The prior odds in our case is simply 1.
The log Bayes factors listed in Table \ref{tab:shank_bf} suggest generally strong evidence for a declining selective pressure after progression, with one notable exception for patient 5.
We also provide a Bayes factor that summarizes the joint evidence for $\omega_{1} < \omega_{2}$, which suggest an overall support in favor of the immune relaxation hypothesis (log BF = 2.14), in accordance with previous findings suggesting a general decrease in non-synonymous divergence at late-stage infection \citep{williamson2005, Lemey2007}.

@@

In a second application of the epoch model, we focus on discrete diffusion processes to infer spatio-temporal history from viral gene sequences.
This type of phylogeographic inference, where the sampling locations are considered as discrete geographic traits, has gained popularity in recent years, at least partly because of a flexible and efficient Bayesian implementation that connects dispersal dynamics to sequence evolution in time-measured phylogenies \citep{Lemey2009}.
Recently \citet{Bahl2011} have applied this Bayesian inference framework to investigate the circulation dynamics of global influenza A H3N2 through time.
Since the authors were interested in capturing the heterogeneity in these dynamics over successive seasonal epidemics between 2003 and 2006, they consider discrete traits that are the product of sampling location and sampling time (epidemic season).
% \myedit{svsixteen}{
This means for example that samples collected from October 2002 to February 2003 in New York are assigned to the discrete state labelled as ``New York 2003'', while sampling in the sample location during another seasons will be represented by another state \citep{Bahl2011}.
% }
Not only does this discretization by sampling time seem counterintuitive for a model that emits discrete outcomes as a continuous function of time, it also considerably increases the dimensionality of the CTMC rate matrix and thus the number of parameters to inform by the sparse spatial data.

### 2015-02-24 chapters/chapter06.tex
preparing manuscript for printing

\myedit{svtwenty}{
Specifically we parametrize each epoch specific, possibly multidimensional parameter $\Theta_{i},\; i=1,\ldots,M$ as a log-linear function of $P$ predictors $\mathbf{X}=\left(\mathbf{x_1},\ldots,\mathbf{x_P} \right)$:
}

>>>>

% \myedit{svtwenty}{
Specifically we parametrize each epoch specific, possibly multidimensional parameter $\Theta_{i},\; i=1,\ldots,M$ as a log-linear function of $P$ predictors $\mathbf{X}=\left(\mathbf{x_1},\ldots,\mathbf{x_P} \right)$:
% }

### 2015-02-24 frontback/summary.tex
preparing manuscript for printing

In Chapters~\ref{chap:spread}~and~\ref{chap:pibuss} are published published manuscripts in which I introduce two software packages written and maintained as part of my PhD training.
Specifically, Chapter~\ref{chap:spread} presents software for visualizing phylogeographic reconstructions resulting from Bayesian inference of viral diffusion processes called SPREAD.
This particular program has proven to be quite popular, which is why in Section~\ref{sec:spread_exten} I discuss extensions and improvements which are planned to appear in the second version of the package.
A detailed tutorial on using the current version of this package can be found in Appendix~\ref{app:spread_tuto}.

>>>>

In Chapters~\ref{chap:spread}~and~\ref{chap:pibuss} are published manuscripts in which I introduce two software packages written and maintained as part of my PhD training.
Specifically, Chapter~\ref{chap:spread} presents software for visualizing phylogeographic reconstructions resulting from Bayesian inference of viral diffusion processes called SPREAD.
This particular program has proven to be quite popular, which is why in Section~\ref{sec:spread_exten} I discuss extensions and improvements which are planned to appear in the second version of the package.
A detailed tutorial on using the current version of this package can be found in Appendix~\ref{app:spread_tuto}.
